---
title: "My First Flexdashboard"
output: 
  flexdashboard::flex_dashboard:
    theme:
       bootswatch: lux
    logo: tsu_logo_yellow.png
    orientation: rows
    vertical_layout: scroll
---

```{r setup, include=FALSE}
library(flexdashboard)

```

```{r data_input}

## libraries
library(dplyr)
library(tidyr)
library(phsopendata)    # easily access open data from PHS
library(ggplot2)
library(lubridate)      # parse dates correctly
library(scales)        #  create % sign in value box
library(plotly)
library(RcppRoll)     # for rolling sum function
library(zoo)

## one way to get data from the PHS opendata repository is by the resource id, found in the metadata section

# daily positives by LA
daily_la <- get_resource("427f9a25-db22-4014-a3bc-893b68243055",
                         col_select = c("Date", "CA", "CAName", "DailyPositive")) %>% 
  mutate(Year = as.numeric(substr(Date, 1, 4)))

# LA populations
la_pops <- get_resource("09ebfefb-33f4-4f6a-8312-2d14e2b02ace",
                        col_select = c("Year", "CA", "Sex", "AllAges")) %>% 
  filter(Year %in% c(2020, 2021),
         Sex == "All") %>%  
# need to create 2022 pops. carry down from previous.
 
  complete(CA, Year = 2020:2022) %>% 
  fill(AllAges)


# Join population data to daily_la
daily_la <- left_join(daily_la, la_pops, by = c("CA", "Year")) %>%
   select(-c(Sex, Year)) %>% 
  rename(Population = AllAges)

# daily & cumulative totals
daily_totals <- get_resource("287fc645-4352-4477-9c8c-55bc054b7e76")

# vaccine data 
vac_totals <- get_resource("42f17a3c-a4db-4965-ba68-3dffe6bca13a")

# cumulative figures by deprivation
dep_cases <- get_resource("307f6666-eb35-4587-b2d3-697eb69368be")
dep_hosp <- get_resource("bef9fce8-62b7-4d34-b54d-4581bfbb64f9")

# define my own colour palette
mycols <- c("#00838F", "#FEED58", "#FEB200", "#FA8C00", "#F3511E", "#C52828")


## function to create heatmap by year
heatmap_function <- function(year_pl) { 
  plot <-
    daily_la %>% 
     # change Date to proper Date format
    mutate(Date = as_date(as.character(Date, "%Y-%m-%d"))) %>%
    
    # create rows for date/LA combinations where no entries have been made
    complete(Date, CAName) %>% 
    
    # group to LA for each date
    group_by(CAName) %>%
    mutate(Positive7Day = roll_sum(DailyPositive, 7, align = "right", fill = NA)) %>% 
    ungroup() %>% 
    group_by(Date, CAName) %>% 
    summarise(Positive7Day = sum(Positive7Day),
              Population = sum(Population),
              CrudeRate7DayPositive = Positive7Day/Population * 100000) %>%
    mutate(CrudeRate7DayPositive = replace_na(CrudeRate7DayPositive, 0)) %>%
    mutate(year = lubridate::year(Date)) %>% 
    
    # filter for current year
    filter(year == year_pl) %>% 
    # create plot
    ggplot(., aes(x = Date, y = CAName, fill = CrudeRate7DayPositive)) +
    geom_tile() +
    
     # apply a defined colour palette instead of default
    scale_fill_gradientn(colours = mycols, name  = "case rate per 100,000") +
    
    # reverse y-axis so in alphabetical order starting at the top
    scale_y_discrete(limits = rev) +
    labs(title = paste0("Heat map of ", year_pl, " case rate per 100,000 for Covid-19"),
         x = year_pl, 
         y = "Local Authority")  +
    theme(legend.position="bottom")+
    guides(fill = guide_legend(label.position = "bottom"))
  plot
}

heat_map_20 <- heatmap_function(2020)
heat_map_21 <- heatmap_function(2021)
heat_map_22 <- heatmap_function(2022)

```


## Introduction {.sidebar}

You could use this space to add information about data sources, how to use the dashboard, assumptions etc.


*Make this text italic *

**Make this text bold **

Data source: 
[PHS Open Data](https://www.opendata.nhs.scot/organization/health_protection) 

Contact: [Sally.Thompson37@nhs.net](mailto:sally.thompson37@nhs.net)\
Last updated: `r format(Sys.time(), '%d %B %Y')` 


### Further Resources

more flexdashboard links here <font size = "1">
<https://bookdown.org/yihui/rmarkdown/dashboards.html>
<https://pkgs.rstudio.com/flexdashboard/> 
</font>


row top 
-----------------------------------------------------------------------

### Total Covid cases in Scotland

```{r}
cum_cases <- daily_totals %>% 
  mutate(Date = as_date(as.character(Date, "%Y-%m-%d"))) %>% 

  filter(Date == max(Date)) %>% 
  pull(CumulativeCases)
  
valueBox(value = prettyNum(cum_cases, big.mark = " "), 
         icon = "fa-virus", 
         caption = "Total Covid-19 cases in Scotland", 
         color = "#ec6555")

```

### 75 and overs who have had 4 vaccines

```{r}
vac_value <- vac_totals %>% 
  mutate(Date = as_date(as.character(Date, "%Y-%m-%d"))) %>% 
           
  filter(Date == max(Date),
         Product == "Total",
         Dose == "Dose 4",
         AgeBand == "75+ years") %>% 
  
  pull(CumulativePercentCoverage) %>% 
  round(., 1)

gauge(vac_value, min = 0, max = 100, symbol = "%") 
```

row mid {.tabset .tabset-fade}
-----------------------------------------------------------------------

### 2022

```{r}

ggplotly(heat_map_22)

```

### 2021

```{r}

ggplotly(heat_map_21)

```

### 2020

```{r}

ggplotly(heat_map_20)

```

## row bottom

### Scotland trend

```{r}

scot_trend <- daily_la %>% 
  
  # change Date to proper Date format
  mutate(Date = as_date(as.character(Date, "%Y-%m-%d")),
         
         # daily trends look too jittery, smooth them by grouping to weekly.  ###### Not summing properly ######
         week = floor_date(Date, "week")) %>%
  
  # need to aggregate all LAs to Scotland level, sum the daily positive and daily cumulative (just in case we want to use later)
  group_by(week) %>% 
  summarise(WeeklyPositive = sum(DailyPositive)
            ) %>% 
  ungroup() %>% 
  
   # filter out last week as it may not include full 7 day totals
  filter(week != max(week)) %>% 
  
  # create time series plot
  ggplot(., aes(x = week, y = WeeklyPositive)) +
  geom_line() +
  
  labs(title = "Weekly cases of Covid-19 across Scotland",
       x = "Date",
       y = "Number of cases each week")+
  
  theme_minimal()
 
ggplotly(scot_trend)
```

### Cases and Hospitalisation by deprivation

```{r}
dep_cases <- dep_cases %>% 
  left_join(., dep_hosp) %>% 
  filter(SIMDQuintile %in% c("1", "2", "3", "4", "5"))

dep_plot <- ggplot(dep_cases, aes(x = TotalCases, y = NumberAdmitted, colour = SIMDQuintile)) +
  geom_point(size = 3) +
  expand_limits(x = 0, y = 0) +
  
  labs(title = "Covid-19 cases and hospitalisations by deprivation quintile",
       x = "Number testing positive for Covid-19",
       y = "Hospitalisations")+
  
  theme_minimal()

ggplotly(dep_plot)
```